Quantifying the tape of life: Ancestry-based metrics provide insights and intuition about evolutionary dynamics

### 2018-04-03 paper.tex
Update on Overleaf.

% *****************
%  TODOs:
% *****************
% [ ] Writing: Abstract
% [ ] Writing: Results & discussion
% [ ] Writing: Conclusions
% [ ] Big 'ol citation pass: need to work appropriate citations into paper

@@

% Alternative titles:
%  - A Suite of Metrics for Understanding Evolutionary Dynamics/Histories
\title{Understanding fine temporal scale evolutionary dynamics with lineage metrics}
\author{Emily Dolson$^{1,2,3}$, Alexander Lalejini$^{1,2, 3}$, Steven Jorgensen$^{1,2}$ \and Charles Ofria$^{1,2, 3}$ \\
\mbox{}\\
$^1$BEACON Center for the Study of Evolution in Action, Michigan State University, East Lansing, MI, 48824 \\
$^2$Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, 48824 \\
$^3$Ecology, Evolutionary Biology, and Behavior Program, Michigan State University, East Lansing, MI, 48824 \\
dolsonem@msu.edu} % email of corresponding author

@@

\begin{abstract}
Fine-scale evolutionary dynamics can be challenging to tease out when focused on broad brush stokes of whole populations over long timespans.
We propose a suite of diagnostic metrics that operate on lineages and phylogenies in digital evolution experiments with the aim of improving our capacity to quantitatively explore the nuances of evolutionary histories in digital evolution experiments.
We present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility. 
Additionally, we suggest the adoption of four phylogeny measurements from biology: depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity. 
We demonstrate the use of each metric on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths, confirming our intuitions about what they can tell us about evolutionary dynamics.

@@

Evolution is a collective effect of many smaller events such as replication, variation, and competition that occur on a fine-grained temporal scale. While evolution's emergent nature %is a large part of why it is so 
can be
fascinating, it also presents challenges to studying the short-term mechanisms that, in aggregate, govern long-term results. In computational evolutionary systems, we can theoretically collect data to help untangle these mechanisms. In practice, however, %gathering this data can be challenging. 
%Moreover, because evolutionary effects can be witnessed on a vastly different scale from their causes, the quantity of such data can be overwhelming.
the sheer number of constituent events produce an overwhelming quantity of data.
%Both of these problems can be mitigated with
In response, we have developed
a standardized set of diagnostic metrics that summarize short-term evolutionary dynamics within a population. Here, we describe a suite of such metrics that operate on lineages and phylogenies, and provide experimental results to help develop an intuition for what they can tell us about evolution.

A lineage describes a continuous line of descent, linking %ancestors and their descendants. 
parents and offspring in an unbroken chain from an original ancestor.
A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation. Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for 
%fine-grained lineage tracking at resolutions that are 
perfect lineage tracking at a level of granularity that is
impossible in modern wet lab experiments.  These data allow us to replay the tape of life in exacting detail and to tease apart the evolutionary recipe for whatever complex feature or phenomenon we happen to be interested in~\citep{mcphee_using_2016}. 
In 
%a more famous
one notable
example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}. %@SPJ: 'More famous example' sounds a bit weird to me

Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize or interpret~\citep{mcphee_visualizing_2016}.
Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual's lineage. The question is how to effectively summarize %what is essentially 
a path through fitness space. One useful abstraction is to treat the path as a sequence of states. Here, we use phenotypes and genotypes as the states in the sequence, but we could just as easily use some other descriptor of the lineage's position in the fitness landscape at a given point in time.
With this abstraction in hand, a few metrics are easily formalized: the number of unique states, the number of transitions between states, and the amount of time spent in each state.
Additionally, we may care about how the transitions between states happened. What mutations led to them? Were those mutations beneficial, deleterious, or neutral at the time? These mutations are particularly notable because they did not simply appear briefly, but stood the test of time, leaving descendants in the final population.
%@CAO: Does the sentence I added above make sense?  My point being that these aren't just a random sampling of possible mutations that we are considering.  It's almost worth an extra sentence to point out that any deleterious mutations on the lineage are of special interest since they made it into the lineage despite being actively selected against.
% @AML: I think so. My interpretation: all mutations (beneficial, deleterious, neutral, etc., etc.) that appear on a lineage of an extant individual are particularly notable. 
%Here, we choose a subset of these metrics that we expect will be broadly useful to explore further.
%@ELD: I agree. It's also nice to emphasize in light of deleterious mutations being the most interesting part of our results.
Here, we explore a subset of these metrics that we expect will be broadly useful.

Whereas a lineage recounts the evolutionary history of a single individual, a phylogeny details the evolutionary history of an entire population. Measurements that summarize phylogenies can provide useful insight into population-level evolutionary dynamics, such as diversification and co-existence among different clades. A variety of useful phylogeny measurements have already been developed by biologists~\citep{tucker_guide_2017}. These measurements tend to treat the phylogeny as a graph and make calculations about its topology. Tucker \textit{et al.} group them into three broad categories: assessments of the quantity of evolutionary history represented by a population, assessments of the amount of divergence within that evolutionary history, and assessments of the topological regularity of the phylogenetic tree. Such measurements can help quantify the behavior of the population as a whole, providing insight into interactions between its members. Thus, they are useful indicators of the presence of various types of eco-evolutionary dynamics.
%@AML: I think the next bit (just moved it) might belong in the metrics section. Instead, we might want a little more elaboration on each of the categories of metrics.
%@ELD: Is that too handwavy?
%@AML: Nah! Handwaving is just what we needed!

%Here, we present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility. Additionally, we suggest the adoption of four phylogeny measurements from biology: depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
Here, we present three types of lineage measurements and suggest adopting four phylogeny measurements from biology; these are lineage length, mutation accumulation, phenotypic volatility, depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
For each metric, we discuss its application and our intuition for what it can tell us about evolution. We evaluate our intuition on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths. For this work, we restrict our attention to asexually reproducing populations; however, we suggest how these metrics can extend to sexual populations.

%@AML:  It might be nice to wrap this section up with a little bit of a 'call to action'/'why should you care about this as a community'
%@ELD: I am in favor.
In addition to demonstrating a range of metrics that are immediately applicable and useful to digital evolution research, we intend for this work to begin a conversation within the %computational evolution
artificial life
community about how we quantify, interpret, and compare observed evolutionary histories. There have been extensive efforts to improve our ability to visualize both lineages and phylogenies~\citep{standish_visualising_2002,burlacu_visualization_2013, mcphee_using_2016,mcphee_visualizing_2016, lalejini_evolutionary_2016}, which are %incredibly valuable 
indispensable 
for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history. However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-based metrics for computational evolution. %Indeed, .... % people have done stuff with lineages
% @CAO: For the moment I commented the "Indeed" out since I'm not sure where it's going; think the current version is a fine place to stop, though I also agree that some of the points below would be good to include if we can.
% - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
% - well-defined metrics not only provide valuable tools for teasing apart evolutionary dynamics, they also facilitate allow us to move away from exclusively qualitative descriptions/understandings of results ==> better quantitative understanding/statistics ==> better hypothesis testing
% - facilitate comparisons among systems. 
% - Something about activity metrics?

@@

Code for all of our metrics is open source and available in the Empirical library (\url{https://github.com/devosoft/Empirical}). Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software. To simplify including useful features with minimal overhead, Empirical is header-only, meaning these metrics can be imported into an existing project with limited effort.
% [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
%@ELD: They should just need the systematics manager. And they might need to hook up a couple of data collection signals.

\subsection{Lineage Metrics}
Each of the three lineage metrics that we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or sequence of individuals that share a common genotypic or phenotypic characteristic of interest; Figure \ref{fig:lineage_metrics_cartoon} is given as a toy example to help guide our discussion of these metrics. 
While we limit our focus to three lineage metrics, this abstraction places lineages in a form suitable for a wide range of measurements, including the direct application of many data mining techniques designed to operate over sequences such as sequential pattern mining, trend analysis, \textit{et cetera} \citep{han2011data}.

@@

\begin{figure*}[t]
\includegraphics[width=\textwidth]{figs/lineage_metrics_cartoon.png}
\caption{\small Panels (A), (B), (C), and (D) each show different ways of representing a lineage. The lineage has accumulated three mutations: one reverse mutation and two substitutions. The phenotype of ancestors changed three times. In (A), each ancestral state along the lineage represents a single individual. Here, the lineage length is equal to the number of generations spanned by the lineage (eight). In (B), ancestral states represent the sequence of genotypes along the lineage. Here, the lineage length is equal to four. In (C) ancestral states represent the sequence of phenotypes along the lineage. Here, the lineage length is equal to the number of times a different phenotype is expressed along the lineage (three). In (D), ancestral states are given by a particular phenotypic characteristic. Here, lineage length is equal to two.}
\label{fig:lineage_metrics_cartoon}
\end{figure*}

\subsubsection{Lineage Length}
Lineage length describes the number of ancestral states traversed by a lineage. If a state represents a single individual along a lineage, lineage length is a measure of the number of generations on a lineage. This form of the lineage length metric is most useful in systems where generational turnover is not algorithmically determined, but instead determined by the life history strategies of organisms. In these systems, lineage length can provide a proxy for replication rate where, for lineages that span equal lengths of time, longer lineages imply faster replication rates (\textit{e.g.} r-selected lineage) and shorter lineages imply slower replication rates (\textit{e.g.} K-selected lineage). 
%@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
%@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.
%@ELD: I like the idea of putting it in that context. I was just suggesting we replace "replicator" with "selected"
%@AML: Put the parentheticals back.

Lineage length becomes an increasingly flexible and informative metric if we begin to consider more abstract definitions of an ancestral state along a lineage. 
We might measure lineage length where an ancestral state represents a temporally continuous sequence of individuals that share a phenotypic or genotypic characteristic of interest. In these cases, lineage length only increases if the characteristic of interest changes from parent to offspring. For example, in an environment where organisms must perform a set of tasks to be successful, we might define ancestral state to be the set of tasks performed by an individual (\textit{e.g.} the set of logic operations performed by a digital organism in the default logic 9 Avida environment). In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes; any sequential ancestors that perform the identical sets of tasks would be compressed into a single state in the sequence. 
% This could be made very clear with a good example here. 
%@ELD Maybe I'm just getting tired, but I didn't quite follow that last sentence
%@AML: It was definitely hard to follow. Hopefully it's better now?
%@ELD: That definitely helped. I tried tweaking it a little more (I think the term "ancestral state" was confusing me, because I'm used to it being the state of the ancestor at the root of the tree)
%  -- Still could use a sentence about what the measure might tell us in the given example.

\subsubsection{Mutation Accumulation}
Mutation accumulation defines a set of measurements that track mutational changes across a lineage. These changes can be measured in a variety of ways, such as the magnitude of the change (for real-valued genomes), or the total count of changes (for discrete-valued genomes). The type of a mutation can also be tracked, and can provide insight into the distribution of mutation types for a given lineage. Measures of mutation accumulation along the lineages of successful extant individuals can help to tease apart the relative importance of different types of mutational events when compared to what is expected by chance.

In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked. Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.  
% @AML: the below points likely have many things that could be cited to support. 
The number and magnitude of deleterious mutations along a lineage can both tell us something about the ruggedness of the fitness landscape, as well as how well a lineage was able to cross fitness valleys~\citep{covert_experiments_2013}. 
Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around. 
%@CAO There are various other things it could tell us, such as additional co-evolutionary pressures that are not reflected in the raw calculation of fitness.  Interestingly, using these metrics in combinaton with local fitness landscape analyses would do the trick.  But that's for another paper.  Indeed, I think it would be nice to put out a series of papers of useful analysis techniques and then do a master paper that reviews them all and demonstrates them in the context of an Avida-like system AND a very different digital evolution system (Markov Brains?)

@@

In scenarios where selective sweeps happen rapidly, the dominant lineage will likely be similar to the average lineage,
%in the population,
as most of the population will be closely related. When the population contains more phylogenetic diversity, however, it is possible for the dominant lineage to differ from the mean. Of course, the nature of such differences is likely informative about the evolutionary dynamics occurring in the population. 
%Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
% @AML: Not sure what the above sentence is saying.
%Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population. % @SPJ: I am not sure what argument this is referring to? The fact that there may or may not be a dominant lineage that differs from the mean?
%@ELD: The point was just that sometimes the dominant lineage is different from the mean and that probably matters but we're not going to talk about it. Since we're going to be short on space, I think we can just cut these lines.

@@

These metrics operate on entire phylogenies rather than single lineages within a population.
%Thus, in most cases, summarizing them across the population is not an issue as it is with lineage metrics.
Thus, there is no need to identify a representative organism or lineage.
Because they %factor in 
use date from
the entire population, phylogeny metrics can be more computationally expensive to calculate than metrics about a single lineage. On the other hand, because most lineages in the population share substantial history, phylogeny metrics can usually be calculated more rapidly than lineage metrics %can be calculated across
for an entire population. Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc. Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.

A standard technique for saving memory and time when working with phylogenties in computational systems is to ``prune" them, removing 
%any taxa that are not either currently alive or the ancestor of taxa that are alive.
% either dead (extinct) branches, or ancestors common to the entire population.
%@CAO: (NEVERMIND - SIMPLIFYING FURTHER... Does the substition I made above make sense?  I'm not quite sure what was meant by "the ancestor of taxa that are alive".
dead (extinct) branches.
Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies. Thus, for the remainder of this paper, we will assume we are working with pruned phylogenies.

@@

An important distinction between phylogenies in natural versus computational systems to be aware of is that natural phylogenies are generally inferred from extant taxa, whereas computational phylogenies are directly recorded. The inferred phylogenies used in natural systems do not contain internal nodes except at branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all organisms in the extant population. For consistency, we will exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.

@@

The depth of the MRCA %(i.e. how many steps from the beginning of time did it occur?) 
(i.e., the number of steps it is from the original ancestor)
%is a useful metric because of the relatively high ratio of informativeness to ease of calculation.
is an informative metric and is easy to calculate.
A recent MRCA implies frequent selective sweeps and less long-term stable coexistence between clades. %The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already. 
Measuring the frequency with which the MRCA changes (i.e. the number of coalescence events) can also be informative, as some conditions can inflate the length of the lineage relative to other conditions without actually increasing the frequency of selective sweeps. This scenario is particularly likely when the population size is changing over time. A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never change its MRCA after the very beginning of evolution %(which is informative as a way of detecting stably coexisting branches in the phylogeny).
(which at least allows us to detect stably coexistance in the population).

@@

Measurements of phylogenetic richness quantify the total amount of evolutionary history contained in a set of taxa. The most traditional metric of phylogenetic richness is the one simply titled ``Phylogenetic Diversity", which is calculated as the number of nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}. Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}. A third approach is to sum evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, across all extant taxa~\citep{tucker_guide_2017}.

@@

Measurements of phylogenetic regularity quantify how balanced the branches of the phylogeny are and are often the variances of values that can be calculated for individual taxa. Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity. The same is true of the variance of evolutionary distinctiveness across the population.

>>>>

\newcommand{\reduceCaptionPadding}{\setlength{\belowcaptionskip}{-10pt}}

\renewcommand\floatpagefraction{.9}
\renewcommand\dblfloatpagefraction{.9} % for two column documents
\renewcommand\topfraction{.9}
\renewcommand\dbltopfraction{.9} % for two column documents
\renewcommand\bottomfraction{.9}
\renewcommand\textfraction{.1}

@@

% Alternative titles:
%  - A Suite of Metrics for Understanding Evolutionary Dynamics/Histories
%  - Quantifying Evolutionary Histories 
%  - A Suite of Metrics for Quantifying Evolutionary Histories
\title{Quantifying the tape of life: Ancestry-based metrics provide insights and intuition about evolutionary dynamics}
\author{Emily Dolson$^{1,2,3}$, Alexander Lalejini$^{1,2, 3}$, Steven Jorgensen$^{1,2}$ \and Charles Ofria$^{1,2, 3}$ \\
\mbox{}\\
$^1$BEACON Center for the Study of Evolution in Action, Michigan State University, East Lansing, MI, 48824 \\
$^2$Department of Computer Science and Engineering, Michigan State University, East Lansing, MI, 48824 \\
$^3$Ecology, Evolutionary Biology, and Behavior Program, Michigan State University, East Lansing, MI, 48824 \\
dolsonem@msu.edu} % email of corresponding author

@@

\begin{abstract}
Fine-scale evolutionary dynamics can be challenging to tease out when focused on broad brush strokes of whole populations over long timespans.
We propose a suite of diagnostic metrics that operate on lineages and phylogenies in digital evolution experiments with the aim of improving our capacity to quantitatively explore the nuances of evolutionary histories in digital evolution experiments.
We present three types of lineage measurements: lineage length, mutation accumulation, and phenotypic volatility. 
Additionally, we suggest the adoption of four phylogeny measurements from biology: depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity. 
We demonstrate the use of each metric on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths, confirming our intuitions about what they can tell us about evolutionary dynamics.

@@

Evolution is a collective effect of many smaller events such as replication, variation, and competition that occur on a fine-grained temporal scale. While evolution's emergent nature can be fascinating, it also presents challenges to studying the short-term mechanisms that, in aggregate, govern long-term results. In computational evolutionary systems, we can theoretically collect data to help untangle these mechanisms. In practice, however, the sheer number of constituent events produce an overwhelming quantity of data.
In response, we have developed a standardized suite of diagnostic metrics to summarize short-term evolutionary dynamics within a population by measuring lineages and phylogenies. Here, we describe these metrics and provide experimental results to develop an intuition for what they can tell us about evolution.

A lineage describes a continuous line of descent, linking parents and offspring in an unbroken chain from an original ancestor.
A complete lineage can provide a post-hoc, step-by-step guide to the evolution of an extant organism where each step involves replication and inherited variation. Indeed, lineage analyses are a powerful tool for disentangling evolutionary dynamics in both natural and digital systems; digital systems, however, allow for perfect lineage tracking at a level of granularity that is impossible in modern wet lab experiments.  These data allow us to replay the tape of life in precise detail and to tease apart the evolutionary recipe for any phenomenon we are interested in~\citep{mcphee_using_2016}. 
In one notable example, Lenski \textit{et al.}, used the lineage of an evolved digital organism in Avida to tease apart, step-by-step how a complex feature (the capacity to perform the equals logical operation) emerged~\citep{lenski_evolutionary_2003}.

Yet, tracking the full details of a single lineage, much less a population of lineages, can be computationally expensive and will inevitably generate an unwieldy amount of data that can be challenging to visualize or interpret~\citep{mcphee_visualizing_2016}.
Summary statistics can help alleviate these issues by enabling the user to focus on aggregate trends across a population rather than needing to examine each individual's lineage. The question is how to effectively summarize a path through fitness space. One useful abstraction is to treat the path as a sequence of states. Here, we use phenotypes and genotypes as the states in the sequence, but we could just as easily use some other descriptor of the lineage's position in the fitness landscape at a given point in time.
With this abstraction in hand, a few metrics are easily formalized: the number of unique states, the number of transitions between states, and the amount of time spent in each state.
Additionally, we may care about how the transitions between states happened. What mutations led to them? Were those mutations beneficial, deleterious, or neutral at the time? These mutations are particularly notable because they did not simply appear briefly, but stood the test of time, leaving descendants in the final population.
%@CAO: Does the sentence I added above make sense?  My point being that these aren't just a random sampling of possible mutations that we are considering.  It's almost worth an extra sentence to point out that any deleterious mutations on the lineage are of special interest since they made it into the lineage despite being actively selected against.
% @AML: I think so. My interpretation: all mutations (beneficial, deleterious, neutral, etc., etc.) that appear on a lineage of an extant individual are particularly notable. 
%Here, we choose a subset of these metrics that we expect will be broadly useful to explore further.
%@ELD: I agree. It's also nice to emphasize in light of deleterious mutations being the most interesting part of our results.
Here, we explore a subset of these metrics that we expect will be broadly useful.

Whereas a lineage recounts the evolutionary history of a single individual, a phylogeny details the evolutionary history of an entire population. Measurements that summarize phylogenies can provide useful insight into population-level evolutionary dynamics, such as diversification and co-existence among different clades. A variety of useful phylogeny measurements have already been developed by biologists~\citep{tucker_guide_2017}. These measurements tend to treat the phylogeny as a graph and make calculations about its topology. Tucker \textit{et al.} group them into three broad categories: assessments of the quantity of evolutionary history represented by a population, assessments of the amount of divergence within that evolutionary history, and assessments of the topological regularity of the phylogenetic tree. Such measurements can help quantify the behavior of the population as a whole, providing insight into interactions between its members. Thus, they are useful indicators of the presence of various types of eco-evolutionary dynamics.

Here, we present three types of lineage measurements and suggest adopting four phylogeny measurements from biology; these are lineage length, mutation accumulation, phenotypic volatility, depth of the most-recent common ancestor, phylogenetic richness, phylogenetic divergence, and phylogenetic regularity.
For each metric, we discuss its application and our intuition for what it can tell us about evolution. We evaluate our intuition on a set of two-dimensional, real-valued optimization problems under a range of mutation rates and selection strengths. For this work, we restrict our attention to asexually reproducing populations; however, we suggest how these metrics can extend to sexual populations.

In addition to demonstrating a range of metrics that are %immediately applicable and
useful to digital evolution research, we intend for this work to begin a conversation within the %computational evolution
artificial life community about how we quantify, interpret, and compare observed evolutionary histories. There have been extensive efforts to improve our ability to represent and visualize both lineages and phylogenies~\citep{standish_visualising_2002,burlacu_visualization_2013, mcphee_using_2016,mcphee_visualizing_2016, lalejini_evolutionary_2016}, which are indispensable for building intuitions and qualitatively understanding the dynamics embedded in a population's evolutionary history. However, we are unaware of efforts to formalize a suite of quantitative lineage and phylogeny-based metrics for computational evolution. %Indeed, .... % people have done stuff with lineages
% @CAO: For the moment I commented the "Indeed" out since I'm not sure where it's going; think the current version is a fine place to stop, though I also agree that some of the points below would be good to include if we can.
% - often a component of a larger experiment-specific analysis or limited to qualitative descriptions/visualizations
% - well-defined metrics not only provide valuable tools for teasing apart evolutionary dynamics, they also facilitate allow us to move away from exclusively qualitative descriptions/understandings of results ==> better quantitative understanding/statistics ==> better hypothesis testing
% - facilitate comparisons among systems. 
% - Something about activity metrics?

@@

Code for all of our metrics is open source and available in the Empirical library (\url{https://github.com/devosoft/Empirical}). Empirical is a C++ library built to facilitate writing efficient and easily sharable scientific software. %To simplify including useful features with minimal overhead, 
Empirical is a header-only library, so adding these metrics to an existing project has minimal overhead.
% [QUESTION] @AML: Is it actually true that someone could include these metrics with minimal effort? How much of Empirical would it impose on the user? Just the systematics manager? 
%@ELD: They should just need the systematics manager. And they might need to hook up a couple of data collection signals.

\subsection{Lineage Metrics}

\begin{figure*}[t]
\includegraphics[width=\textwidth]{figs/lineage_metrics_cartoon.png}
\reduceCaptionPadding
\caption{\small Four methods of representing a lineage.  This example lineage has accumulated three mutations (one reverse mutation and two substitutions) and gone through three distinct phenotypes. In (A), each state along the lineage represents a single individual; lineage length is the number of generations spanned by the lineage (eight). In (B), states represent the sequence of genotypes along the lineage, reducing lineage length to four. In (C) states represent the sequence of phenotypes along the lineage; lineage length is the number of times a different phenotype is expressed (three). In (D), states are a particular phenotypic characteristic; here, lineage length is two.}
\label{fig:lineage_metrics_cartoon}
\end{figure*}

Each of the three lineage metrics that we discuss --- lineage length, mutation accumulation, and phenotypic volatility --- reduces a lineage to a linear sequence of states where each state represents an individual or sequence of individuals that share a common genotypic or phenotypic characteristic of interest; Figure \ref{fig:lineage_metrics_cartoon} is given as a toy example to help guide our discussion of these metrics. 
While we limit our focus to three lineage metrics, this abstraction places lineages in a form suitable for a wide range of measurements, including the direct application of many data mining techniques designed to operate over sequences such as sequential pattern mining, trend analysis, \textit{et cetera} \citep{han2011data}.

@@

\subsubsection{Lineage Length}
Lineage length describes the number of states traversed by a lineage. If a state is defined as a single individual, %along a lineage,
lineage length is a count of the number of generations.
%on a lineage.
%This form of the lineage length metric 
Generation count is most useful in systems where generational turnover is not fixed, %algorithmically determined,
but instead determined by the life history strategies of organisms.
%In these systems, %lineage length can provide a proxy for replication rate where, 
For lineages that span equal lengths of time, more generations imply faster replication rates (\textit{e.g.} r-selected lineage) while fewer generations imply slower replication rates (\textit{e.g.} K-selected lineage). 
%@ELD: I've always heard these called r- and k- selected rather than replicator, but I'm totally willing to believe these terms are used too.
%@AML: Yup. I have no idea if I've ever heard them referred to as 'r-replicator vs. k-replicator'. I'm going to get rid of the two parentheticals for now.
%@ELD: I like the idea of putting it in that context. I was just suggesting we replace "replicator" with "selected"
%@AML: Put the parentheticals back.

Lineage length becomes %an increasingly 
a more flexible and informative metric if we consider more abstract definitions of %an ancestral 
states along a lineage. 
We might measure lineage length where a state represents a %temporally continuous 
sequence of individuals that share a particular phenotypic or genotypic characteristic.
%of interest. 
In these cases, lineage length only increases when the characteristic of interest changes from parent to offspring. For example, in an environment where organisms must perform tasks to be successful, we might define state as the set of tasks performed by an individual. %(\textit{e.g.} the set of logic operations performed by a digital organism in the default logic-9 Avida environment).
In this scenario, lineage length would only increase when the set of tasks performed by an ancestor changes; sequential ancestors that perform the identical sets of tasks would be compressed into a single state in the sequence, even if other traits differ. 
% This could be made very clear with a good example here. 
%@ELD Maybe I'm just getting tired, but I didn't quite follow that last sentence
%@AML: It was definitely hard to follow. Hopefully it's better now?
%@ELD: That definitely helped. I tried tweaking it a little more (I think the term "ancestral state" was confusing me, because I'm used to it being the state of the ancestor at the root of the tree)
%  -- Still could use a sentence about what the measure might tell us in the given example.

\subsubsection{Mutation Accumulation}
Mutation accumulation defines a set of measurements that track mutational changes across a lineage. These changes can be measured %in a variety of ways, such 
as the magnitude of the change (for real-valued genomes) or as the total count of changes (for discrete-valued genomes). Mutation effects can also be tracked to gain insights about their distribution along a given lineage. Measures of mutation accumulation along the lineages of successful individuals can help tease apart the relative importance of different types of mutational events when compared to what is expected by chance.

In conjunction with collected fitness information, the class of a mutation (\textit{e.g.} beneficial, deleterious, or neutral) can also be tracked. Different evolutionary conditions are expected to cause different distributions of mutations along a lineage~\citep{barrick2013genome}; deviations revealed by measures of mutation accumulation can act as a barometer for unexpected evolutionary dynamics.  
% @AML: the below points likely have many things that could be cited to support. 
The number and magnitude of deleterious mutations along a lineage can tell us both about the ruggedness of the fitness landscape, and about a lineage's ability to cross fitness valleys~\citep{covert_experiments_2013}. 
Similarly, an elevated measure of neutral mutations relative to beneficial or deleterious mutations can tell us that the fitness landscape has a large neutral area that the lineage is spending most of its time drifting around. 
%@CAO There are various other things it could tell us, such as additional co-evolutionary pressures that are not reflected in the raw calculation of fitness.  Interestingly, using these metrics in combinaton with local fitness landscape analyses would do the trick.  But that's for another paper.  Indeed, I think it would be nice to put out a series of papers of useful analysis techniques and then do a master paper that reviews them all and demonstrates them in the context of an Avida-like system AND a very different digital evolution system (Markov Brains?)

@@

In scenarios with frequent selective sweeps, the dominant lineage will likely be similar to the average lineage,
%in the population,
as most of the population will be closely related. When the population contains more phylogenetic diversity, however, %it is possible for 
the dominant lineage may differ from the mean. Of course, the nature of such differences is likely informative about the evolutionary dynamics occurring in the population. 
%Since we have yet to observe an instance of this phenomenon, we do not explore it further in this paper. 
% @AML: Not sure what the above sentence is saying.
%Nevertheless, it presents an argument for measuring the dominant lineage in addition to gathering summary statistics across the population. % @SPJ: I am not sure what argument this is referring to? The fact that there may or may not be a dominant lineage that differs from the mean?
%@ELD: The point was just that sometimes the dominant lineage is different from the mean and that probably matters but we're not going to talk about it. Since we're going to be short on space, I think we can just cut these lines.

@@

These metrics operate on entire phylogenies rather than single lineages within a population, eliminating the need to identify a representative organism or lineage.
Because they use data from the entire population, phylogeny metrics can be more computationally expensive to calculate than single lineage metrics. On the other hand, because most lineages tend to share substantial history, phylogeny metrics can usually be calculated more rapidly than full-population lineage metrics. %can be calculated across an entire population.
Note that phylogenies can be constructed with regard to any taxonomic level of organization, be it individual, genotype, phenotype, etc. Thus, when we refer generally to items in a phylogeny, we will use the term \textit{taxa}.

A standard technique for saving memory and time when working with phylogenies in computational systems is to ``prune" them, removing 
%any taxa that are not either currently alive or the ancestor of taxa that are alive.
% either dead (extinct) branches, or ancestors common to the entire population.
%@CAO: (NEVERMIND - SIMPLIFYING FURTHER... Does the substition I made above make sense?  I'm not quite sure what was meant by "the ancestor of taxa that are alive".
dead (extinct) branches.
Since all of the phylogeny metrics we discuss here are borrowed from natural systems (where we do not have information about taxa without offspring), they all are designed to work on pruned phylogenies. Thus, for the remainder of this paper, we will assume we are working with pruned phylogenies.

@@

An important distinction between phylogenies in natural versus computational systems 
%to be aware of 
is that natural phylogenies are generally inferred from extant taxa, whereas computational phylogenies are directly recorded. %The inferred phylogenies used in natural systems 
Inferred phylogenies 
do not contain internal nodes except at branch points. They also do not contain history prior to the most recent common ancestor (MRCA) of all extant organisms. %in the extant population.
For consistency, we exclude pre-MRCA taxa from our analyses. However, we will not remove non-branching internal nodes, as these only serve to make our phylogenies more informative.

@@

The depth of the MRCA %(i.e. how many steps from the beginning of time did it occur?) 
(i.e., the number of steps it is from the original ancestor)
%is a useful metric because of the relatively high ratio of informativeness to ease of calculation.
is an informative metric and is easy to calculate.
A recent MRCA implies frequent selective sweeps and less long-term stable coexistence between clades. %The depth of the MRCA is easily calculated because most implementations of phylogeny pruning require keeping track of it already. 
Measuring the frequency with which the MRCA changes (i.e. the number of coalescence events) can also be informative, as some conditions can inflate the length of the lineage relative to other conditions without actually increasing the frequency of selective sweeps. This scenario is particularly likely when the population size is changing over time. A downside to the depth of MRCA as a metric is that any population that does have a stable ecology will likely never change its MRCA after the very beginning of evolution %(which is informative as a way of detecting stably coexisting branches in the phylogeny).
(which at least allows us to detect stably coexistence in the population).

@@

Measurements of phylogenetic richness quantify the total amount of evolutionary history contained in a set of taxa. The most traditional metric of phylogenetic richness is
%the one simply titled
``Phylogenetic Diversity", which is calculated as the number of nodes in the minimum spanning tree from the MRCA to all extant taxa~\citep{faith_conservation_1992}. Another approach is to calculate the pairwise distances between all taxa and sum them~\citep{tucker_guide_2017}. A third approach is to sum evolutionary distinctiveness, a measurement of a taxon's evolutionary uniqueness~\citep{isaac_mammals_2007}, across all extant taxa~\citep{tucker_guide_2017}.

@@

Measurements of phylogenetic regularity quantify how balanced the branches are in a phylogeny and are often the variances of values calculated for individual taxa. Just as the mean of the pairwise distances between all taxa in the population is a measurement of phylogenetic divergence, taking their variance is a measurement of phylogenetic regularity. The same is true of the variance of evolutionary distinctiveness across the population.
